Fast Fourier transform of electromagnetic data for computationally expensive kernels

### 2021-05-25 ManuscriptTime.tex
Add link to original article

\title[Transforming expensive EM kernels]{Fast Fourier transformation
of electromagnetic data for computationally expensive kernels}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data is often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of such modelling that is obtained through a transform is defined by
three key points: solver, method and implementation of the Fourier transform,
and gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points results in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.
CSEM codes for 2D and 3D computations, on the other hand, often compute their
responses directly in the required domain, either frequency or time.

@@

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity of different time-domain methods with the
computational complexity of frequency-domain computation followed by a Fourier
transform: explicit time-stepping schemes (Du Fort-Frankel method as used in,
e.g., \cite{GEO.04.Commer, GEO.07.Maao}) and implicit schemes \citep[e.g.,
][]{GEO.04.Haber,GEO.10.Um} have a complexity of $O(n^4)$, matrix exponentials
and Lanczos reduction schemes \cite[e.g., ][]{RS.94.Druskin} have a complexity
of $O(n^4\sqrt{\log n})$, and Fourier transform of frequency-domain solutions
have a complexity $O(n_f n^3)$; $n$ is the number of grid points in each
direction, and $n_f$ the number of frequencies. Their conclusion was that the
Fourier transform method can be favourable if the number of required
frequencies is small relative to the number of grid points in each coordinate.
The conditions for fast computation of time-domain data with a frequency-domain
code are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are the further reduction of required frequencies
by introducing lower and upper thresholds of numerically important frequencies,
an adaptive, frequency-dependent gridding scheme that minimizes the required
cells in each dimension, and a logarithmic Fourier transform such as digital
linear filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast Fourier
transform \citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency to the
time domain. The latter makes it also possible to only use the imaginary part
of the frequency-domain response, which has advantages when it comes to
interpolation.

In the next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of the methodology, our
changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results. These
compromise of a homogeneous space, a layered model, a model which includes
dispersive media (induced polarization), and a three-dimensional model.

@@

The proposed Fourier transform requires a robust solver that can compute
accurate results over a wide range of frequencies (times) to obtain time-domain
(frequency-domain) responses. In our case the solver uses a regular, stretched
grid and computes the electric fields in the frequency domain. This setup is
the target of the following gridding recommendations. Whilst they will look
differently for other mesh types or a time-domain code, some conclusion will
still hold.

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the boundary
condition. It can also have a much higher stretching factor $\alpha_\mr{c}$. In
the presented examples we have chosen $D_\mr{c}$ such that the distance for the
signal diffusing from the source to the boundary and back to the receiver
closest to the boundary is at least two wavelengths, after which the amplitude
of the signal is reduced to a millionth of its initial strength. The wavelength
$\lambda$ to compute $D_\mr{c}$ is given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth. This also
applies to the horizontal dimensions, not only to the upward $z$-direction, and
it equally applies to very resistive basements, even in deep water. One way to
circumvent this difficulty is the use of a primary-secondary formulation, where
the primary field, including the air wave, is computed with a semi-analytical
code for layered media. We do not consider this approach here. Note that grid
stretching for complex-valued diffusion fields is essentially what is done for
wave fields with the perfectly matched layer (PML). PML is an absorbing layer
to avoid scattering from the boundary by letting the field decay to zero, which
is achieved by introducing a complex factor that causes damping. As
electromagnetic diffusion fields are damped fields by themselves, it suffices
to stretch the grid.

@@

\cite{GJI.17.Commer} have shown that time-domain IP modelling can be difficult,
particularly for low $c$-values. Modelling frequency-dependent models in the
frequency domain is, for any given frequency, not different from modelling a
non frequency-dependent model. However, the interesting point is to see how the
Fourier transform behaves to obtain time-domain data from frequency-domain
responses. As an example, we modelled the response for a land case with air
above a halfspace of 1\,S/m. Within the halfspace is a dispersive, 100\,m thick
layer at a depth of 300\,m, with $\sigma_0=1.0\,$S/m and
$\sigma_\infty=1.25\,$S/m; $\tau=1$, and we vary $c$ from 0 (no IP) to 1.0,
0.75, 0.5, and 0.25. The x-directed step-off source and the receiver at an
inline offset of 900\,m are place at the surface. The real and imaginary
frequency-domain responses for this model are shown in the left column of
Figure~\ref{fig:CCM}. It can be seen that the effect of the IP layer in the
frequency domain is that the real part starts to deviate from the DC value at
lower frequencies. The imaginary part decays slower towards zero when
decreasing the frequency, and it deviates from a power-law function at lower
frequencies as well. This indicates that we need lower frequency content for
the Fourier transform the smaller $c$ becomes.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{11-cole-cole-model}
  \caption{IP example for different $c$ values, frequency-domain response on
    the left and time-domain response on the right. The frequencies indicated
    by the blue vertical bars in the lower-left figure were used for our
    Fourier transform approach, requiring only the computation of 27
    frequencies instead of the full 747 frequencies required by the digital
    linear filter method.}
  \label{fig:CCM}
\end{figure*}
%

>>>>

\title[Transforming expensive EM kernels]{Fast Fourier transform
of electromagnetic data for computationally expensive kernels}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data are often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of modelling obtained through such a transform is defined by three
key points: solver, method and implementation of the Fourier transform, and
gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points resulted in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.
CSEM codes for arbitrary 2D and 3D computations, on the other hand, often
compute their responses directly in the required domain, either frequency or
time.

@@

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity of different time-domain methods with the
computational complexity of frequency-domain computation followed by a Fourier
transform: explicit time-stepping schemes (Du Fort-Frankel method as used in,
e.g., \cite{GEO.04.Commer, GEO.07.Maao}) and implicit schemes \citep[e.g.,
][]{GEO.04.Haber,GEO.10.Um} have a complexity of $O(n^4)$, matrix exponentials
and Lanczos reduction schemes \cite[e.g., ][]{RS.94.Druskin} have a complexity
of $O(n^4\sqrt{\log n})$, and Fourier transforms of frequency-domain solutions
have a complexity $O(n_f n^3)$; $n$ is the number of grid points in each
direction, and $n_f$ the number of frequencies. Their conclusion was that the
Fourier transform method can be favourable if the number of required
frequencies is small relative to the number of grid points in each coordinate.
The conditions for fast computation of time-domain data with a frequency-domain
code are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are the further reduction of required frequencies
by introducing lower and upper thresholds of numerically important frequencies,
an adaptive, frequency-dependent gridding scheme that minimizes the required
cells in each dimension, and a logarithmic Fourier transform such as digital
linear filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast Fourier
transform \citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency to the
time domain. The latter makes it also possible to only use the imaginary part
of the frequency-domain response, which has advantages when it comes to
interpolation.

In the next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of the methodology, our
changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results. These
comprise in of a homogeneous space, a layered model, a model which includes
dispersive media (induced polarization), and a three-dimensional model.

@@

The proposed Fourier transform requires a robust solver that can compute
accurate results over a wide range of frequencies (times) to obtain time-domain
(frequency-domain) responses. In our case the solver uses a regular, stretched
grid and computes the electric fields in the frequency domain. This setup is
the target of the following gridding recommendations. Whilst they will look
differently for other mesh types or a time-domain code, some conclusions will
still hold.

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
a very small cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the boundary
condition. It can also have a much higher stretching factor $\alpha_\mr{c}$. In
the presented examples we have chosen $D_\mr{c}$ such that the distance for the
signal diffusing from the source to the boundary and back to the receiver
closest to the boundary is at least two wavelengths, after which the amplitude
of the signal is reduced to a millionth of its initial strength. The wavelength
$\lambda$ to compute $D_\mr{c}$ is given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth. This also
applies to the horizontal dimensions, not only to the upward $z$-direction, and
it equally applies to very resistive basements, even in deep water. One way to
circumvent this difficulty is the use of a primary-secondary formulation, where
the primary field, including the air wave, is computed with a semi-analytical
code for layered media. We do not consider this approach here. Note that grid
stretching for complex-valued diffusion fields is essentially what is done for
wave fields with the perfectly matched layer (PML). PML is an absorbing layer
to avoid scattering from the boundary by letting the field decay to zero, which
is achieved by introducing a complex factor that causes damping. As
electromagnetic diffusion fields are damped fields by themselves, it suffices
to stretch the grid.

@@

\cite{GJI.17.Commer} have shown that time-domain IP modelling can be difficult,
particularly for low $c$-values. Modelling frequency-dependent models in the
frequency domain is, for any given frequency, not different from modelling a
non frequency-dependent model. However, the interesting point is to see how the
Fourier transform behaves to obtain time-domain data from frequency-domain
responses. As an example, we modelled the response for a land case with air
above a halfspace of 1\,S/m. Within the halfspace is a dispersive, 100\,m thick
layer at a depth of 300\,m, with $\sigma_0=1.0\,$S/m and
$\sigma_\infty=1.25\,$S/m; $\tau=1$, and we vary $c$ from 0 (no IP) to 1.0,
0.75, 0.5, and 0.25. The x-directed step-off source and the receiver at an
inline offset of 900\,m are placed on the surface. The real and imaginary
frequency-domain responses for this model are shown in the left column of
Figure~\ref{fig:CCM}. It can be seen that the effect of the IP layer in the
frequency domain is that the real part starts to deviate from the DC value at
lower frequencies. The imaginary part decays slower towards zero when
decreasing the frequency, and it deviates from a power-law function at lower
frequencies as well. This indicates that we need lower frequency content for
the Fourier transform, the smaller $c$ becomes.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{11-cole-cole-model}
  \caption{IP example for different $c$ values, frequency-domain response on
    the left and time-domain response on the right. The frequencies indicated
    by the blue vertical bars in the lower-left figure were used for our
    Fourier transform approach, requiring only the computation of 27
    frequencies instead of the full 747 frequencies required by the digital
    linear filter method.}
  \label{fig:CCM}
\end{figure*}
%
